---
title: "Getting Started with openbiologging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with openbiologging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

## Overview

The `openbiologging` package provides data from a systematic literature review
of open data practices in biologging research. Biologging uses miniaturized
animal-attached sensors to record movement, behavior, physiology, and
environment. This review assessed how openly researchers share the data
collected from these devices.

The package contains two datasets:

- **`biologging`** — one row per reviewed paper, covering 4,799 publications
  identified from Web of Science Core Collection (queried 2023-08-21)
- **`taxa`** — one row per species per study, for studies where more than one
  species was tagged

```{r setup}
library(openbiologging)
library(dplyr)
library(ggplot2)
```

## The `biologging` dataset

```{r biologging-overview}
glimpse(biologging)
```

Each row is a paper. The key columns fall into a few groups:

**What kind of paper is it?**

- `manuscript_type` — Study (`S`), Review (`R`), Method (`M`), Perspective
  (`P`), or Data paper (`D`)
- `novel_biologging` — `Y` if the paper collected new biologging data; `N` if
  it reanalysed existing data

**What was tracked, and where?**

- `biologging_context` — Wild (`W`), Captive (`C`), Domesticated (`D`), or
  combinations
- `habitat` — Aquatic (`A`), Marine (`M`), Terrestrial (`T`), or combinations
- `device_cat` — sensor categories used: Location (`L`), Intrinsic (`I`),
  Environment (`E`), or combinations
- `genus` / `species` — focal taxon (primary species for multi-species studies)

**Was the data shared?**

- `biologging_availability` — `Y` if the paper includes a data availability
  statement specifically addressing the biologging data; `N` otherwise
- `external_data` — `Y` if the paper incorporated non-biologging data from
  external sources (e.g. satellite imagery, phylogenies)

## The `taxa` dataset

```{r taxa-overview}
glimpse(taxa)
```

`taxa` covers multi-species studies — papers where `more_species == "Y"` in
`biologging`. Join the two datasets on `id` to get full taxonomic information
alongside the review scores:

```{r join-example, warning=FALSE, message=FALSE}
taxa |>
  left_join(biologging, by = c("id", "genus", "species")) |>
  select(id, manuscript_type, biologging_context, device_cat, habitat, biologging_availability, genus, species, title, authors, doi) |>
  slice_head(n = 10)
```

## Exploring the data

### How open is biologging data sharing?

Do biologging papers include a data availability statement for their data?

```{r availability, warning=FALSE, message=FALSE}
biologging |>
  filter(!is.na(biologging_availability)) |>
  count(biologging_availability) |>
  mutate(
    label = if_else(biologging_availability == "Y", "Available", "Not available"),
    pct   = round(n / sum(n) * 100, 1)
  )
```

### Manuscript types

Most papers in the review are empirical studies:

```{r manuscript-type, warning=FALSE, message=FALSE}
biologging |>
  count(manuscript_type, sort = TRUE) |>
  mutate(
    type_label = case_match(
      manuscript_type,
      "S" ~ "Study",
      "R" ~ "Review",
      "M" ~ "Method",
      "P" ~ "Perspective",
      "D" ~ "Data paper"
    ),
    pct = round(n / sum(n) * 100, 1)
  )
```

### Habitat representation over time

Single-habitat papers (A, M, or T only) broken down by year:

```{r habitat-over-time, warning=FALSE, message=FALSE}
biologging |>
  filter(!is.na(year), !is.na(habitat), habitat %in% c("A", "M", "T")) |>
  count(year, habitat) |>
  ggplot(aes(x = factor(year), y = n, fill = habitat)) +
  geom_col() +
  scale_y_continuous(breaks = c(0, 150, 300, 450, 600)) +
  scale_fill_manual(
    values = c("A" = "#1b9e77", "M" = "#d95f02", "T" = "#7570b3"),
    labels = c("A" = "Aquatic", "M" = "Marine", "T" = "Terrestrial"),
    name   = "Habitat"
  ) +
  labs(y = "Number of papers") +
  theme_minimal() +
  theme(
    axis.title.x  = element_blank(),
    axis.title.y  = element_text(margin = margin(10, 10, 10, 10)),
    axis.text.x   = element_text(angle = 45, hjust = 1),
    legend.box.margin = margin(t = 20)
  )
```

### Sensor categories

The `device_cat` column uses single letters for sensor types and combines them
for multi-sensor tags. The most common combinations:

```{r device-cat, warning=FALSE, message=FALSE}
biologging |>
  filter(!is.na(device_cat)) |>
  count(device_cat, sort = TRUE) |>
  slice_head(n = 8) |>
  mutate(
    device_label = case_match(
      device_cat,
      "L"   ~ "Location only",
      "LI"  ~ "Location + Intrinsic",
      "LIE" ~ "Location + Intrinsic + Environment",
      "LE"  ~ "Location + Environment",
      "I"   ~ "Intrinsic only",
      "IE"  ~ "Intrinsic + Environment",
      "E"   ~ "Environment only",
      "LI"  ~ "Location + Intrinsic",
      .default = device_cat
    )
  ) |>
  ggplot(aes(x = reorder(device_label, n), y = n)) +
  geom_col(fill = "#7570b3") +
  coord_flip() +
  labs(x = NULL, y = "Number of papers") +
  theme_minimal()
```

### Biologging context

Were animals tracked in wild, captive, or domesticated settings?

```{r context, warning=FALSE, message=FALSE}
biologging |>
  filter(novel_biologging == "TRUE", !is.na(biologging_context)) |>
  count(biologging_context, sort = TRUE) |>
  mutate(
    context_label = case_match(
      biologging_context,
      "W"  ~ "Wild",
      "C"  ~ "Captive",
      "D"  ~ "Domesticated",
      "WC" ~ "Wild + Captive",
      "WD" ~ "Wild + Domesticated",
      "CD" ~ "Captive + Domesticated",
      .default = biologging_context
    ),
    pct = round(n / sum(n) * 100, 1)
  ) |>
  select(context_label, n, pct)
```

### Most studied taxa

Which genera appear most frequently across all studies? This uses both
`biologging` (single-species studies) and `taxa` (multi-species studies):

```{r top-taxa, warning=FALSE, message=FALSE}
all_taxa <- bind_rows(
  biologging |> select(genus, species),
  taxa       |> select(genus, species)
)

all_taxa |>
  filter(!is.na(genus)) |>
  count(genus, sort = TRUE) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(genus, n), y = n)) +
  geom_col(fill = "#d95f02") +
  coord_flip() +
  labs(x = NULL, y = "Number of studies") +
  theme_minimal()
```

## Further reading

- Full methods: <https://flukeandfeather.github.io/openbiologging/methods.html>
- Sensor category definitions: Williams et al. (2019),
  doi:[10.1111/1365-2656.13094](https://doi.org/10.1111/1365-2656.13094)
- Package documentation: `?biologging`, `?taxa`
